---
title: "accessibility"
format: html
editor: visual
---

## 1. Load data

```{r}
library(tidyverse)
library(sf)

recreational_parks <- st_read(
  "clean_data/recreational_parks.gpkg"
)

facility_level <- read_csv(
  "clean_data/facility_level.csv"
)

sport_level <- read_csv(
  "clean_data/sport_level.csv"
)

lga <- st_read(
  "data/lga/LGA_POLYGON.shp"
)
```

## 2. Prepare LGA boundaries

```{r}
# Transform recreational spaces to the LGA coordinate reference system
recreational_parks <- recreational_parks |>
  st_transform(st_crs(lga))
```

```{r}
# Create a list of the 79 incorporated Victorian LGAs
lga_master <- lga |>
  st_drop_geometry() |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  select(LGA_CODE, LGA_NAME) |>
  distinct() |>
  arrange(LGA_NAME)

nrow(lga_master)
```

## 3. Calculate park accessibility

### 3.1 Assign parks to LGAs for park counts

For park-count calculations, each recreational space is assigned to the LGA
containing its largest overlapping area. This prevents parks crossing LGA
boundaries from being counted more than once.

```{r, warning=FALSE}
parks_lga_count <- st_join(
  recreational_parks,
  lga |>
    select(LGA_CODE, LGA_NAME),
  left = TRUE,
  largest = TRUE
)
```

### 3.2 Inspect unmatched parks

```{r}
parks_lga_count |>
  filter(is.na(LGA_CODE)) |>
  st_drop_geometry() |>
  select(PRIMS_ID, NAME, AREA_TYPE)
```

```{r}
parks_lga_count <- parks_lga_count |>
  filter(!is.na(LGA_CODE))
```

### 3.3 Calculate park counts by LGA

```{r}
park_count_lga <- parks_lga_count |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_count = n_distinct(PRIMS_ID),
    .groups = "drop"
  )
```

### 3.4 Intersect parks with LGA boundaries

Park polygons are intersected with LGA boundaries so that only the park area
located within each LGA is included in that LGA's total park area.

```{r, warning=FALSE, message=FALSE}
parks_lga_area <- st_intersection(
  recreational_parks |>
    select(PRIMS_ID),
  lga |>
    select(LGA_CODE, LGA_NAME)
)
```

```{r}
parks_lga_area <- parks_lga_area |>
  mutate(
    intersection_area_sqm =
      as.numeric(st_area(parks_lga_area))
  )
```

### 3.5 Calculate park area by LGA

```{r}
park_area_lga <- parks_lga_area |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_area_sqm =
      sum(intersection_area_sqm, na.rm = TRUE),

    park_area_hectares =
      park_area_sqm / 10000,

    .groups = "drop"
  )
```

### 3.6 Combine park count and area indicators

```{r}
park_accessibility_lga <- lga_master |>
  left_join(
    park_count_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    park_area_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  mutate(
    park_count = replace_na(park_count, 0L),
    park_area_sqm = replace_na(park_area_sqm, 0),
    park_area_hectares =
      replace_na(park_area_hectares, 0)
  )
```

```{r}
park_accessibility_lga |>
  summarise(
    n_lgas = n(),
    total_parks = sum(park_count),
    total_park_area_hectares =
      sum(park_area_hectares)
  )
```

## 4. Calculate facility accessibility

### 4.1 facility level

```{r}
facility_accessibility_lga <- facility_level |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    facility_count =
      n_distinct(facility_id),

    total_fields_courts =
      sum(n_fields_courts, na.rm = TRUE),

    mean_sports_per_facility =
      mean(n_sports, na.rm = TRUE),

    .groups = "drop"
  )
```

### 4.2 sport level

```{r}
sport_accessibility_lga <- sport_level |>
  filter(
    !str_detect(LGA_NAME, "UNINC"),
    !is.na(sports_played)
  ) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    sport_opportunity_count =
      n_distinct(facilty_sport_played_id),

    sport_type_count =
      n_distinct(sports_played),

    facilities_with_sport_data =
      n_distinct(facility_id),

    .groups = "drop"
  )
```

## 5. Combine accessibility indicators


```{r}
lga_master <- lga_master |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

park_accessibility_lga <- park_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

facility_accessibility_lga <- facility_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

sport_accessibility_lga <- sport_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )
```


```{r}
accessibility_raw_lga <- lga_master |>
  left_join(
    park_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    facility_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    sport_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  )
```

### Check the combined dataset

```{r}
accessibility_raw_lga |>
  summarise(
    n_rows = n(),
    n_lga_codes = n_distinct(LGA_CODE),
    n_lga_names = n_distinct(LGA_NAME)
  )
```

```{r}
accessibility_raw_lga |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.x))
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  )
```

