---
title: "accessibility"
format: html
editor: visual
---

# Construct LGA-level accessibility indicators

## Load data

```{r}
library(tidyverse)
library(sf)
library(readxl)
library(corrplot)

recreational_parks <- st_read(
  "clean_data/recreational_parks.gpkg"
)

facility_level <- read_csv(
  "clean_data/facility_level.csv"
)

sport_level <- read_csv(
  "clean_data/sport_level.csv"
)

lga <- st_read(
  "data/lga/LGA_POLYGON.shp"
)

population_lga_raw <- read_excel(
  "data/Population_ LGA2022-2023.xlsx",
  sheet = "Table 2",
  skip = 7,
  col_names = c(
    "ABS_LGA_CODE",
    "LGA_NAME",
    "population_2022",
    "population_2023",
    "population_change",
    "population_change_percent",
    "natural_increase",
    "net_internal_migration",
    "net_overseas_migration",
    "area_km2",
    "population_density_2023"
  )
)
```

## Prepare LGA boundaries

```{r}
# Transform recreational spaces to the LGA coordinate reference system
recreational_parks <- recreational_parks |>
  st_transform(st_crs(lga))
```

```{r}
# Create a list of the 79 incorporated Victorian LGAs
lga_master <- lga |>
  st_drop_geometry() |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  select(LGA_CODE, LGA_NAME) |>
  distinct() |>
  arrange(LGA_NAME)

nrow(lga_master)
```

## Prepare LGA population 

```{r}
population_lga_raw <- population_lga_raw |>
  filter(
    !is.na(ABS_LGA_CODE),
    LGA_NAME != "Unincorporated Vic"
  ) |>
  mutate(
    ABS_LGA_CODE = as.character(ABS_LGA_CODE),

    LGA_NAME = LGA_NAME |>
      str_remove(" \\(Vic\\.\\)$") |>
      str_to_upper(),

    population_2022 = as.numeric(population_2022),
    population_2023 = as.numeric(population_2023),

    area_km2 = as.numeric(area_km2),
    population_density_2023 =
      as.numeric(population_density_2023)
  )

# Keep 2023 LGA population
population_lga <- population_lga_raw |>
  select(
    LGA_NAME,
    ABS_LGA_CODE,
    population_2023,
    area_km2,
    population_density_2023
  )
```

## Calculate park accessibility

### Assign parks to LGAs for park counts

For park-count calculations, each recreational park is counted in every incorporated LGA that it intersects. Therefore, a park crossing an LGA boundary may contribute to the park count of more than one LGA.

```{r}
parks_lga_count <- st_join(
  recreational_parks,
  lga |>
    select(LGA_CODE, LGA_NAME),
  join = st_intersects,
  left = FALSE
)
```

### Inspect LGAs with no park intersections

```{r}
# Find LGAs with no intersecting park polygons
lga_master |>
  anti_join(
    parks_lga_count |>
      st_drop_geometry() |>
      filter(!str_detect(LGA_NAME, "UNINC")) |>
      distinct(LGA_NAME),
    by = "LGA_NAME"
  )
```


### Calculate park counts by LGA

```{r}
park_count_lga <- parks_lga_count |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_count = n_distinct(PRIMS_ID),
    .groups = "drop"
  ) |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )
```

### Intersect parks with LGA boundaries

Park polygons are intersected with LGA boundaries so that only the park area
located within each LGA is included in that LGA's total park area.

```{r, warning=FALSE, message=FALSE}
parks_lga_area <- st_intersection(
  recreational_parks |>
    select(PRIMS_ID),
  lga |>
    select(LGA_CODE, LGA_NAME)
)
```

```{r}
parks_lga_area <- parks_lga_area |>
  mutate(
    intersection_area_sqm =
      as.numeric(st_area(parks_lga_area))
  )
```

### Calculate park area by LGA

```{r}
park_area_lga <- parks_lga_area |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_area_sqm =
      sum(intersection_area_sqm, na.rm = TRUE),
    .groups = "drop"
  )
```

### Combine park count and area indicators

```{r}
park_accessibility_lga <- lga_master |>
  left_join(
    park_count_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    park_area_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  mutate(
    park_count = replace_na(park_count, 0L),
    park_area_sqm = replace_na(park_area_sqm, 0)
  )
```

```{r}
park_accessibility_lga |>
  summarise(
    n_lgas = n(),
    total_parks = sum(park_count)
  )
```

## Calculate facility accessibility

### facility level

```{r}
facility_accessibility_lga <- facility_level |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    facility_count =
      n_distinct(facility_id),

    total_fields_courts =
      sum(n_fields_courts, na.rm = TRUE),

    mean_sports_per_facility =
      mean(n_sports, na.rm = TRUE),

    .groups = "drop"
  )
```

### sport level

```{r}
sport_accessibility_lga <- sport_level |>
  filter(
    !str_detect(LGA_NAME, "UNINC"),
    !is.na(sports_played)
  ) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    sport_opportunity_count =
      n_distinct(facilty_sport_played_id),

    sport_type_count =
      n_distinct(sports_played),

    facilities_with_sport_data =
      n_distinct(facility_id),

    .groups = "drop"
  )
```

## Combine accessibility indicators

```{r}
lga_master <- lga_master |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

park_accessibility_lga <- park_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

facility_accessibility_lga <- facility_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

sport_accessibility_lga <- sport_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )
```


```{r}
accessibility_lga <- lga_master |>
  left_join(
    park_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    facility_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    sport_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  )
```

### Check the combined dataset

```{r}
accessibility_lga |>
  summarise(
    n_rows = n(),
    n_lga_codes = n_distinct(LGA_CODE),
    n_lga_names = n_distinct(LGA_NAME)
  )
```

```{r}
accessibility_lga |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.x))
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  )
```

## Add population data and calculate population-based accessibility

```{r}
accessibility_lga <- accessibility_lga |>
  left_join(
    population_lga |>
      select(
        LGA_NAME,
        population_2023,
        area_km2,
        population_density_2023
      ),
    by = "LGA_NAME"
  )

accessibility_lga <- accessibility_lga |>
  mutate(
    parks_per_10000 =
      park_count /
      population_2023 *
      10000,

    park_area_sqm_per_capita =
      park_area_sqm /
      population_2023,

    facilities_per_10000 =
      facility_count /
      population_2023 *
      10000,

    fields_courts_per_10000 =
      total_fields_courts /
      population_2023 *
      10000,

    sport_opportunities_per_10000 =
      sport_opportunity_count /
      population_2023 *
      10000,

    sport_data_coverage =
      facilities_with_sport_data /
      facility_count
  )
```

- **Park provision**
  - `parks_per_10000`
  - `park_area_sqm_per_capita`

- **Facility provision**
  - `facilities_per_10000`
  - `fields_courts_per_10000`
  - `sport_opportunities_per_10000`

- **Diversity**
  - `sport_type_count`
  - `mean_sports_per_facility`

`facilities_with_sport_data` and `sport_data_coverage` are data quality variables, not accessibility predictors.

```{r}
# Save accessibility dataset
write.csv(
  accessibility_lga,
  "clean_data/accessibility_lga.csv"
)
```

# Validate and prepare accessibility indicators for regression

## Check distributions

```{r}
accessibility_lga |>
  select(
    parks_per_10000,
    park_area_sqm_per_capita,
    facilities_per_10000,
    fields_courts_per_10000,
    sport_opportunities_per_10000,
    sport_type_count,
    mean_sports_per_facility
  ) |>
  summary()

accessibility_lga |>
  select(
    parks_per_10000,
    park_area_sqm_per_capita,
    facilities_per_10000,
    fields_courts_per_10000,
    sport_opportunities_per_10000,
    sport_type_count,
    mean_sports_per_facility
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(
    ~ variable,
    scales = "free"
  ) +
  theme_minimal()
```

Several accessibility indicators were strongly right-skewed. In particular, parks per 10,000 residents and park area per capita had large positive skewness, while facility-related rate measures were also moderately to strongly right-skewed. In contrast, sport type count and mean sports per facility were approximately symmetric.

## Check correlations between accessibility indicators

```{r}
accessibility_cor <- accessibility_lga |>
  select(
    parks_per_10000,
    park_area_sqm_per_capita,
    facilities_per_10000,
    fields_courts_per_10000,
    sport_opportunities_per_10000,
    sport_type_count,
    mean_sports_per_facility
  ) |>
  cor(
    use = "complete.obs",
    method = "spearman"
  ) |>
  round(2)

# correlation matrix
accessibility_cor

# scatterplot matrix
GGally::ggpairs(
  accessibility_lga |>
    select(
      parks_per_10000,
      park_area_sqm_per_capita,
      facilities_per_10000,
      fields_courts_per_10000,
      sport_opportunities_per_10000,
      sport_type_count,
      mean_sports_per_facility
    )
) +
  theme(
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 6)
  )

# correlation heatmap
corrplot(
  accessibility_cor,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.7,
  number.cex = 0.7
)
```

Strong correlations were observed among several accessibility measures. The three facility-related measures were especially highly correlated, with correlations ranging from 0.94 to 0.99. The two park measures were also strongly correlated. These highly correlated variables should not all be included in the same regression model.