---
title: "accessibility"
format: html
editor: visual
---

## Load data

```{r}
library(tidyverse)
library(sf)
library(readxl)

recreational_parks <- st_read(
  "clean_data/recreational_parks.gpkg"
)

facility_level <- read_csv(
  "clean_data/facility_level.csv"
)

sport_level <- read_csv(
  "clean_data/sport_level.csv"
)

lga <- st_read(
  "data/lga/LGA_POLYGON.shp"
)

population_lga_raw <- read_excel(
  "data/Population_ LGA2022-2023.xlsx",
  sheet = "Table 2",
  skip = 7,
  col_names = c(
    "ABS_LGA_CODE",
    "LGA_NAME",
    "population_2022",
    "population_2023",
    "population_change",
    "population_change_percent",
    "natural_increase",
    "net_internal_migration",
    "net_overseas_migration",
    "area_km2",
    "population_density_2023"
  )
)
```

## Prepare LGA boundaries

```{r}
# Transform recreational spaces to the LGA coordinate reference system
recreational_parks <- recreational_parks |>
  st_transform(st_crs(lga))
```

```{r}
# Create a list of the 79 incorporated Victorian LGAs
lga_master <- lga |>
  st_drop_geometry() |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  select(LGA_CODE, LGA_NAME) |>
  distinct() |>
  arrange(LGA_NAME)

nrow(lga_master)
```

## Prepare LGA population 

```{r}
population_lga_raw <- population_lga_raw |>
  filter(
    !is.na(ABS_LGA_CODE),
    LGA_NAME != "Unincorporated Vic"
  ) |>
  mutate(
    ABS_LGA_CODE = as.character(ABS_LGA_CODE),

    LGA_NAME = LGA_NAME |>
      str_remove(" \\(Vic\\.\\)$") |>
      str_to_upper(),

    population_2022 = as.numeric(population_2022),
    population_2023 = as.numeric(population_2023),

    area_km2 = as.numeric(area_km2),
    population_density_2023 =
      as.numeric(population_density_2023)
  )

# keep 2023 LGA population
population_lga <- population_lga_raw |>
  select(
    LGA_NAME,
    ABS_LGA_CODE,
    population_2023,
    area_km2,
    population_density_2023
  )
```

## Calculate park accessibility

### Assign parks to LGAs for park counts

For park-count calculations, each recreational space is assigned to the LGA
containing its largest overlapping area. This prevents parks crossing LGA
boundaries from being counted more than once.

```{r, warning=FALSE}
parks_lga_count <- st_join(
  recreational_parks,
  lga |>
    select(LGA_CODE, LGA_NAME),
  left = TRUE,
  largest = TRUE
)
```

### Inspect unmatched parks

```{r}
parks_lga_count |>
  filter(is.na(LGA_CODE)) |>
  st_drop_geometry() |>
  select(PRIMS_ID, NAME, AREA_TYPE)
```

```{r}
parks_lga_count <- parks_lga_count |>
  filter(!is.na(LGA_CODE))
```

### Calculate park counts by LGA

```{r}
park_count_lga <- parks_lga_count |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_count = n_distinct(PRIMS_ID),
    .groups = "drop"
  )
```

### Intersect parks with LGA boundaries

Park polygons are intersected with LGA boundaries so that only the park area
located within each LGA is included in that LGA's total park area.

```{r, warning=FALSE, message=FALSE}
parks_lga_area <- st_intersection(
  recreational_parks |>
    select(PRIMS_ID),
  lga |>
    select(LGA_CODE, LGA_NAME)
)
```

```{r}
parks_lga_area <- parks_lga_area |>
  mutate(
    intersection_area_sqm =
      as.numeric(st_area(parks_lga_area))
  )
```

### Calculate park area by LGA

```{r}
park_area_lga <- parks_lga_area |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_area_sqm =
      sum(intersection_area_sqm, na.rm = TRUE),

    park_area_hectares =
      park_area_sqm / 10000,

    .groups = "drop"
  )
```

### Combine park count and area indicators

```{r}
park_accessibility_lga <- lga_master |>
  left_join(
    park_count_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    park_area_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  mutate(
    park_count = replace_na(park_count, 0L),
    park_area_sqm = replace_na(park_area_sqm, 0),
    park_area_hectares =
      replace_na(park_area_hectares, 0)
  )
```

```{r}
park_accessibility_lga |>
  summarise(
    n_lgas = n(),
    total_parks = sum(park_count),
    total_park_area_hectares =
      sum(park_area_hectares)
  )
```

## Calculate facility accessibility

### facility level

```{r}
facility_accessibility_lga <- facility_level |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    facility_count =
      n_distinct(facility_id),

    total_fields_courts =
      sum(n_fields_courts, na.rm = TRUE),

    mean_sports_per_facility =
      mean(n_sports, na.rm = TRUE),

    .groups = "drop"
  )
```

### sport level

```{r}
sport_accessibility_lga <- sport_level |>
  filter(
    !str_detect(LGA_NAME, "UNINC"),
    !is.na(sports_played)
  ) |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    sport_opportunity_count =
      n_distinct(facilty_sport_played_id),

    sport_type_count =
      n_distinct(sports_played),

    facilities_with_sport_data =
      n_distinct(facility_id),

    .groups = "drop"
  )
```

## Combine accessibility indicators

```{r}
lga_master <- lga_master |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

park_accessibility_lga <- park_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

facility_accessibility_lga <- facility_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )

sport_accessibility_lga <- sport_accessibility_lga |>
  mutate(
    LGA_CODE = as.character(LGA_CODE)
  )
```


```{r}
accessibility_raw_lga <- lga_master |>
  left_join(
    park_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    facility_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    sport_accessibility_lga,
    by = c("LGA_CODE", "LGA_NAME")
  )
```

### Check the combined dataset

```{r}
accessibility_raw_lga |>
  summarise(
    n_rows = n(),
    n_lga_codes = n_distinct(LGA_CODE),
    n_lga_names = n_distinct(LGA_NAME)
  )
```

```{r}
accessibility_raw_lga |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.x))
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  )
```

