---
title: "Recreational Spaces and Health Outcomes Across Victoria"
format: html
editor: visual
---

```{r}
library(readxl)
library(sf)
library(dplyr)
library(ggplot2)
library(stringr)
```

## Load data

```{r}
# Load the Parks and Conservation Reserves spatial dataset
parks <- st_read("data/parks/PARKRES.shp")

# Load the sport and recreational facilities dataset
facilities <- read_excel("data/facilities.xlsx")

# Load LGA-level physical activity and health data
# n_max is used to exclude non-data rows at the bottom of the worksheet
activity <- read_excel(
  "data/Victorian Population Health Survey 2023.xlsx",
  n_max = 8910
)

# Load Victorian Local Government Area boundaries
lga <- st_read("data/LGA/LGA_POLYGON.shp")
```

## Overview of parks across Victorian LGAs

The following map shows all parks and conservation reserves in the original dataset. It provides an initial overview of their spatial distribution across Victorian LGAs before recreational park types are selected.

```{r, fig.width=8, fig.height=7}
# Map all parks and conservation reserves together with LGA boundaries
ggplot() +
  geom_sf(
    data = parks,
    fill = "#63A65F",
    colour = NA,
    alpha = 0.75
  ) +
  geom_sf(
    data = lga,
    fill = NA,
    colour = "grey30",
    linewidth = 0.25
  ) +
  labs(
    title = str_wrap(
      "Distribution of All Parks and Conservation Reserves Across Victorian LGAs",
      width = 55
    ),
    subtitle = "LGA boundaries are shown in grey",
    caption = paste(
      "Source: Victorian Parks and Conservation",
      "Reserves and LGA boundary datasets"
    )
  ) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold",
      lineheight = 1.1
    ),
    plot.subtitle = element_text(
      size = 11,
      colour = "grey40"
    ),
    plot.caption = element_text(
      size = 9,
      colour = "grey50"
    ),
    plot.margin = margin(15, 20, 15, 20)
  )
```

## Inspect the park data

Before selecting recreational park types, the structure of the original park dataset is examined. In particular, `PRIMS_ID` is checked because it is used as the park identifier.

### Park identifier summary

```{r}
# Summarise the number of polygon records and unique park identifiers
parks |>
  st_drop_geometry() |>
  summarise(
    total_polygon_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID),
    duplicated_id_records = sum(duplicated(PRIMS_ID))
  )
```

```{r}
# Examine how many park identifiers contain one or multiple polygons
parks |>
  st_drop_geometry() |>
  count(PRIMS_ID) |>
  summarise(
    single_polygon_parks = sum(n == 1),
    multi_polygon_parks = sum(n > 1),
    maximum_polygons_per_park = max(n)
  )
```

Multiple rows with the same `PRIMS_ID` do not necessarily represent duplicate parks. A park or reserve may consist of several separate polygon components.

### Check park name consistency

```{r}
# Check whether any park identifier is associated with multiple names
parks |>
  st_drop_geometry() |>
  group_by(PRIMS_ID) |>
  summarise(
    n_names = n_distinct(NAME, na.rm = TRUE),
    names = paste(
      sort(unique(na.omit(NAME))),
      collapse = " | "
    ),
    .groups = "drop"
  ) |>
  filter(n_names > 1)
```

### Check park name and classification consistency

```{r}
# Create a summary of names and park types associated with each park ID
park_consistency <- parks |>
  st_drop_geometry() |>
  group_by(PRIMS_ID) |>
  summarise(
    n_records = n(),
    n_names = n_distinct(NAME, na.rm = TRUE),
    n_area_types = n_distinct(AREA_TYPE, na.rm = TRUE),
    names = paste(
      sort(unique(na.omit(NAME))),
      collapse = " | "
    ),
    area_types = paste(
      sort(unique(na.omit(AREA_TYPE))),
      collapse = " | "
    ),
    .groups = "drop"
  )
```

```{r}
# Display park IDs with inconsistent names or classifications
park_consistency |>
  filter(n_names > 1 | n_area_types > 1) |>
  arrange(desc(n_names), desc(n_area_types)) |>
  print(n = Inf)
```

### Inspect inconsistent park records

```{r}
# Identify park IDs with inconsistent names or classifications
inconsistent_ids <- park_consistency |>
  filter(n_names > 1 | n_area_types > 1) |>
  select(PRIMS_ID)

# Examine the original attributes of these park IDs
parks |>
  st_drop_geometry() |>
  semi_join(inconsistent_ids, by = "PRIMS_ID") |>
  select(
    PRIMS_ID,
    NAME,
    AREA_TYPE,
    MANAGER,
    IUCN,
    HECTARES,
    AREASQM
  ) |>
  distinct() |>
  arrange(PRIMS_ID, NAME, AREA_TYPE) |>
  print(n = Inf)
```

## Inspect park and reserve types

The next step is to examine the available park classifications and identify the types that are publicly accessible and suitable for recreational activities.

```{r}
# Count the number of polygon records in each park classification
parks |>
  st_drop_geometry() |>
  count(AREA_TYPE, sort = TRUE) |>
  print(n = Inf)
```

```{r}
# Count unique parks rather than polygon records in each classification
parks |>
  st_drop_geometry() |>
  distinct(PRIMS_ID, AREA_TYPE) |>
  count(AREA_TYPE, sort = TRUE) |>
  print(n = Inf)
```