---
title: "Recreational Spaces and Health Outcomes Across Victoria"
format: html
editor: visual
---

```{r}
library(readxl)
library(sf)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyverse)
library(readxl)
library(janitor)         
library(skimr) 
```

## Load data

```{r}
# Load the Parks and Conservation Reserves spatial dataset
parks <- st_read("data/parks/PARKRES.shp")

# Load the sport and recreational facilities dataset
facilities <- read_excel("data/facilities.xlsx")

# Load LGA-level physical activity and health data
# n_max is used to exclude non-data rows at the bottom of the worksheet
activity <- read_excel(
  "data/Victorian Population Health Survey 2023.xlsx",
  n_max = 8910
)

# Load Victorian Local Government Area boundaries
lga <- st_read("data/LGA/LGA_POLYGON.shp")
```

## Parks and recreational spaces

### 1. Explore the park data

#### Overview of parks across Victorian LGAs

The following map shows all parks and conservation reserves in the original dataset. It provides an initial overview of their spatial distribution across Victorian LGAs before recreational park types are selected.

```{r, fig.width=8, fig.height=7}
# Map all parks and conservation reserves together with LGA boundaries
ggplot() +
  geom_sf(
    data = parks,
    fill = "#63A65F",
    colour = NA,
    alpha = 0.75
  ) +
  geom_sf(
    data = lga,
    fill = NA,
    colour = "grey30",
    linewidth = 0.25
  ) +
  labs(
    title = str_wrap(
      "Distribution of All Parks and Conservation Reserves Across Victorian LGAs",
      width = 55
    ),
    subtitle = "LGA boundaries are shown in grey",
    caption = paste(
      "Source: Victorian Parks and Conservation",
      "Reserves and LGA boundary datasets"
    )
  ) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold",
      lineheight = 1.1
    ),
    plot.subtitle = element_text(
      size = 11,
      colour = "grey40"
    ),
    plot.caption = element_text(
      size = 9,
      colour = "grey50"
    ),
    plot.margin = margin(15, 20, 15, 20)
  )
```


#### Inspect park data structure

```{r,echo=FALSE}
# Examine the variables and data types in the park dataset
glimpse(parks)
```


#### Check key variables

```{r}
# Check missing values in key park variables
parks |>
  st_drop_geometry() |>
  summarise(
    missing_id = sum(is.na(PRIMS_ID)),
    missing_name = sum(is.na(NAME)),
    missing_area_type = sum(is.na(AREA_TYPE))
  )
```

#### Check park geometries

```{r}
# Check for empty and invalid park geometries
parks |>
  summarise(
    empty_geometries = sum(st_is_empty(geometry)),
    invalid_geometries = sum(!st_is_valid(geometry))
  ) |>
  st_drop_geometry()
```
### 2. Clean and validate park records

#### Check park identifiers

```{r}
# Compare the number of polygon records with unique park identifiers
parks |>
  st_drop_geometry() |>
  summarise(
    polygon_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID),
    additional_polygon_records = n() - n_distinct(PRIMS_ID)
  )
```

#### Check park attribute consistency

```{r}
# Check whether each park ID has consistent names and classifications
parks |>
  st_drop_geometry() |>
  group_by(PRIMS_ID) |>
  summarise(
    n_names = n_distinct(NAME),
    n_area_types = n_distinct(AREA_TYPE),
    .groups = "drop"
  ) |>
  filter(n_names > 1 | n_area_types > 1)
```


##### Inspect inconsistent park attributes

```{r}
# Display the names and classifications of inconsistent park IDs
parks |>
  st_drop_geometry() |>
  filter(PRIMS_ID %in% c(18, 694, 4654, 5077, 5089)) |>
  distinct(PRIMS_ID, NAME, AREA_TYPE) |>
  arrange(PRIMS_ID, NAME) |>
  tibble::as_tibble() |>
  print(n = Inf)
```

##### Examine inconsistent park records

```{r}
# Examine additional attributes of inconsistent park records
parks |>
  st_drop_geometry() |>
  filter(PRIMS_ID %in% c(18, 694, 4654, 5077, 5089)) |>
  select(
    PRIMS_ID,
    NAME,
    AREA_TYPE,
    AREATYPEID,
    ESTAB_DATE,
    LAST_MOD,
    HECTARES
  ) |>
  arrange(PRIMS_ID, NAME, LAST_MOD) |>
  tibble::as_tibble() |>
  print(n = Inf)
```


The checks identified no missing values or invalid geometries. Multiple records
with the same `PRIMS_ID` mainly represent separate components or subdivisions
of the same park, so these records are retained.

### 3. Define recreational spaces


#### Inspect park classifications

```{r}
# Count unique park IDs in each park classification
parks |>
  st_drop_geometry() |>
  distinct(PRIMS_ID, AREA_TYPE) |>
  count(AREA_TYPE, sort = TRUE) |>
  tibble::as_tibble() |>
  print(n = Inf)
```

#### Select recreational park types

```{r}
# Preliminary expanded selection of recreational parks and reserves
recreational_types <- c(
  # Clearly classified parks
  "FOREST PARK",
  "METROPOLITAN PARK",
  "NATIONAL PARK - SCHEDULE 2, NATIONAL PARKS ACT",
  "NATIONAL PARKS ACT SCHEDULE 4 PARK OR RESERVE",
  "OTHER PARK - SCHEDULE 3, NATIONAL PARKS ACT",
  "OTHER PARK - SCHEDULE 9, NATIONAL PARKS ACT",
  "PHILLIP ISLAND NATURE PARK",
  "REGIONAL PARK - NOT SCHEDULED UNDER NATIONAL PARKS ACT",
  "RESERVOIR PARK",
  "STATE PARK - SCHEDULE 2B, NATIONAL PARKS ACT",
  "WILDERNESS PARK - SCHEDULE 2A, NATIONAL PARKS ACT",

  # Reserves likely to support terrestrial outdoor recreation
  "COASTAL RESERVE",
  "NATURAL FEATURES RESERVE",
  "NATURAL FEATURES RESERVE - BUSHLAND RESERVE",
  "NATURAL FEATURES RESERVE - STREAMSIDE RESERVE",
  "NATURAL FEATURES RESERVE - LAKE RESERVE",
  "NATURAL FEATURES RESERVE - SCENIC RESERVE",
  "NATURAL FEATURES RESERVE - NATURAL FEATURES AND SCENIC RESERVE",
  "NATURAL FEATURES RESERVE - GIPPSLAND LAKES RESERVE",
  "NATURAL FEATURES RESERVE - GEOLOGICAL RESERVE",
  "NATURAL FEATURES RESERVE - CAVE RESERVE",
  "NATURAL FEATURES RESERVE - RIVER MURRAY RESERVE"
)

parks_selected <- parks |>
  filter(AREA_TYPE %in% recreational_types)

```

#### Check selected recreational parks

```{r}
# Compare the original and selected park datasets
bind_rows(
  parks |>
    st_drop_geometry() |>
    summarise(
      dataset = "All park records",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    ),
  parks_selected |>
    st_drop_geometry() |>
    summarise(
      dataset = "Selected recreational parks",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    )
)
```


#### Summarise selected park types

```{r}
# Count selected park IDs by classification
parks_selected |>
  st_drop_geometry() |>
  distinct(PRIMS_ID, AREA_TYPE) |>
  count(AREA_TYPE, sort = TRUE) |>
  tibble::as_tibble() |>
  print(n = Inf)
```

### 4. Prepare recreational space polygons


#### Combine park polygons

```{r}
# Combine multiple polygon components belonging to the same park
parks_dissolved <- parks_selected |>
  group_by(PRIMS_ID) |>
  summarise(
    NAME = first(NAME),
    AREA_TYPE = first(AREA_TYPE),
    .groups = "drop"
  )
```

```{r}
# Confirm that each park ID now appears once
parks_dissolved |>
  st_drop_geometry() |>
  summarise(
    park_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID)
  )
```

#### Check coordinate reference systems

```{r}
# Check whether parks and LGA boundaries use the same CRS
st_crs(parks_dissolved) == st_crs(lga)
```


#### Compare original and selected park data

```{r}
# Compare the original dataset with the selected recreational spaces
bind_rows(
  parks |>
    st_drop_geometry() |>
    summarise(
      dataset = "All park records",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    ),
  parks_selected |>
    st_drop_geometry() |>
    summarise(
      dataset = "Selected recreational spaces",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    )
)
```

#### Combine polygons belonging to the same park

A single park may consist of multiple polygon records. These polygons are
combined by `PRIMS_ID` so that each recreational space is counted only once.

```{r}
# Combine multiple polygons belonging to the same park
parks_dissolved <- parks_selected |>
  group_by(PRIMS_ID) |>
  summarise(
    NAME = first(NAME),
    AREA_TYPE = first(AREA_TYPE),
    .groups = "drop"
  )
```

```{r}
# Confirm that there is one record for each park ID
parks_dissolved |>
  st_drop_geometry() |>
  summarise(
    park_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID)
  )
```

### 5. Calculate recreational space accessibility by LGA

#### Assign recreational spaces to LGAs

For park-count calculations, each recreational space is assigned to the LGA
containing its largest overlapping area. This prevents parks crossing LGA
boundaries from being counted more than once.

```{r, warning=FALSE}
# Assign each recreational space to the LGA with the largest overlap
parks_lga_count <- st_join(
  parks_dissolved,
  lga |>
    select(LGA_CODE, LGA_NAME),
  left = TRUE,
  largest = TRUE
)
```

##### Inspect unmatched recreational spaces

```{r}
# Display recreational spaces that were not assigned to an LGA
parks_lga_count |>
  filter(is.na(LGA_NAME)) |>
  st_drop_geometry() |>
  select(PRIMS_ID, NAME, AREA_TYPE)
```

##### Remove recreational spaces outside LGA boundaries

Two coastal reserves do not intersect any LGA boundary and therefore cannot
be included in the LGA-level park-count calculation.

```{r}
# Remove recreational spaces that could not be assigned to an LGA
parks_lga_count <- parks_lga_count |>
  filter(!is.na(LGA_NAME))

# Confirm that all remaining recreational spaces have an assigned LGA
sum(is.na(parks_lga_count$LGA_NAME))
```

##### Calculate recreational space counts by LGA

```{r}
# Count recreational spaces assigned to each LGA
park_count_lga <- parks_lga_count |>
  st_drop_geometry() |>
  count(
    LGA_CODE,
    LGA_NAME,
    name = "park_count"
  ) |>
  arrange(desc(park_count))

park_count_lga |>
  tibble::as_tibble() |>
  print(n = Inf)
```


#### Intersect park recreational spaces with LGA boundaries

Recreational spaces crossing LGA boundaries are divided along those boundaries,
so that only the area located within each LGA is included.

```{r, warning=FALSE}
# Split recreational spaces according to LGA boundaries
parks_lga_area <- st_intersection(
  lga |>
    select(LGA_CODE, LGA_NAME),
  parks_dissolved |>
    select(PRIMS_ID, NAME, AREA_TYPE)
)

# Calculate the area of each intersected section in square metres
parks_lga_area <- parks_lga_area |>
  mutate(
    park_area_sqm = as.numeric(st_area(geometry))
  )
```

```{r}
nrow(parks_lga_area)
```

##### Calculate park recreational space area by LGA

```{r}
# Sum the intersected recreational space area within each LGA
park_area_lga <- parks_lga_area |>
  st_drop_geometry() |>
  group_by(LGA_CODE, LGA_NAME) |>
  summarise(
    park_area_sqm = sum(park_area_sqm, na.rm = TRUE),
    park_area_hectares = park_area_sqm / 10000,
    .groups = "drop"
  ) |>
  arrange(desc(park_area_sqm))

park_area_lga |>
  tibble::as_tibble() |>
  print(n = Inf)
```


#### Combine recreational space counts and areas

```{r}
# Combine park-count and park-area indicators for each LGA
park_accessibility_lga <- lga |>
  st_drop_geometry() |>
  select(LGA_CODE, LGA_NAME) |>
  left_join(
    park_count_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  left_join(
    park_area_lga,
    by = c("LGA_CODE", "LGA_NAME")
  ) |>
  mutate(
    park_count = coalesce(park_count, 0L),
    park_area_sqm = coalesce(park_area_sqm, 0),
    park_area_hectares = coalesce(park_area_hectares, 0)
  ) |>
  arrange(LGA_NAME)

park_accessibility_lga |>
  tibble::as_tibble() |>
  print(n = Inf)
```



## Facility


### 1. Column Name Cleaning
```{r}
facility_df <- facilities %>% 
  clean_names()  # "Facility ID" -> "facility_id"...

glimpse(facility_df)
skim(facility_df)
```

### 2. Preliminary data overview
```{r}
glimpse(facility_df)
skim(facility_df)
dim(facility_df)
n_distinct(facility_df$facility_id_1)
n_distinct(facility_df$facilty_sport_played_id)
```

### 3. Duplicate and record structure checks
```{r}
# Duplicate
facility_df %>%
  duplicated() %>%
  sum()
```

```{r}
# Duplicate facility-sport records
facility_df %>%
  count(
    facility_id_1,
    sports_played,
    sort = TRUE
  ) %>%
  filter(n > 1)
```

```{r}
facility_df %>%
  group_by(facility_id_1, sports_played) %>%
  summarise(
    n_records = n(),
    total_fields = sum(number_of_field_courts, na.rm = TRUE),
    values = paste(
      sort(unique(number_of_field_courts[!is.na(number_of_field_courts)])),
      collapse = ", "
    ),
    .groups = "drop"
  ) %>%
  filter(n_records > 1)
```

Therefore, sport records originate from different components.

```{r}
facility_df %>%
  group_by(facility_id_1) %>%
  summarise(
    n_name = n_distinct(facility_name, na.rm = TRUE),
    n_lga = n_distinct(lga_name, na.rm = TRUE),
    n_lat = n_distinct(latitude, na.rm = TRUE),
    n_lon = n_distinct(longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    n_name > 1 |
    n_lga > 1 |
    n_lat > 1 |
    n_lon > 1
  )

```

```{r}
facility_df %>%
  count(facilty_sport_played_id) %>%
  filter(n > 1)
```

No exact duplicate records were identified. The 9,590 records represent 5,000 unique facilities and multiple facility-sport/component records. Repeated combinations of facility ID and sport were not treated as duplicates because they can correspond to distinct facility-sport records with different sport IDs, playing areas or surface types.

### 4. Convert coordinates & coordinate check
```{r}
# Convert coordinates
facility_df <- facility_df %>%
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )
```

```{r}
# Coordinate validity check
facility_df %>%
  summarise(
    missing_lat = sum(is.na(latitude)),
    missing_lon = sum(is.na(longitude)),
    min_lat = min(latitude, na.rm = TRUE),
    max_lat = max(latitude, na.rm = TRUE),
    min_lon = min(longitude, na.rm = TRUE),
    max_lon = max(longitude, na.rm = TRUE),
    
    invalid_lat = sum(
      !is.na(latitude) &
      (latitude < -39 | latitude > -34)
    ),
    
    invalid_lon = sum(
      !is.na(longitude) &
      (longitude < 140 | longitude > 150)
    )
  )
```

```{r}
facility_df <- facility_df %>%
  mutate(
    coordinate_status = case_when(
      is.na(latitude) | is.na(longitude) ~ "Missing",
      latitude < -39.2 | latitude > -33.98 |
        longitude < 140.96 | longitude > 150.04 ~ "Invalid",
      TRUE ~ "Valid"
    )
  )
```

```{r}
facility_df %>%
  count(coordinate_status)
```

```{r}
# Identify invalid coordinates
facility_df %>%
  filter(coordinate_status == "Invalid") %>%
  select(
    facility_id_1,
    lga_name,
    facility_name,
    suburb_town,
    pcode,
    latitude,
    longitude
  )
```

```{r}
# Resolve/review coordinate errors
facility_df <- facility_df %>%
  mutate(
    latitude_raw = latitude,
    longitude_raw = longitude
  )
```

```{r}
facility_df <- facility_df %>%
  mutate(
    
    latitude = case_when(
      # SWANHI: Swap latitude and longitude
      facility_id_1 %in% c("SWANHI5447", "SWANHI5448", "SWANHI5449") ~ longitude_raw,
      
      # MURRIN: Latitude decimal-place error
      facility_id_1 == "MURRIN5458" ~ -37.494546,
      facility_id_1 == "MURRIN5461" ~ -37.53826,
      
      # BENDIG5468: Outside Victoria, set to missing value.
      facility_id_1 == "BENDIG5468" ~ NA_real_,
      
      # Other cases, the original values ​​are retained.
      TRUE ~ latitude_raw
    ),
    
    longitude = case_when(
      # SWANHI: Swap latitude and longitude
      facility_id_1 %in% c("SWANHI5447", "SWANHI5448", "SWANHI5449") ~ latitude_raw,

      facility_id_1 == "BENDIG5468" ~ NA_real_,
      
      # Other cases, the original values ​​are retained.
      TRUE ~ longitude_raw
    )
  )
```

### 5. Count by Ownership ,Purpose & Category
```{r}
facility_df %>%
  count(facility_ownership, sort = TRUE)

facility_df %>%
  count(facility_purpose, sort = TRUE)

facility_df %>%
  count(facility_category, sort = TRUE)
```

```{r}
facility_df %>%
  group_by(facility_id_1) %>%
  summarise(
    n_ownership = n_distinct(facility_ownership, na.rm = TRUE),
    n_purpose = n_distinct(facility_purpose, na.rm = TRUE),
    n_category = n_distinct(facility_category, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summarise(
    ownership_conflicts = sum(n_ownership > 1),
    purpose_conflicts = sum(n_purpose > 1),
    category_conflicts = sum(n_category > 1)
  )
```

### 6. Source LGA Quality Check

```{r}
facility_df %>%
  count(lga_name, sort = TRUE)

sort(unique(facility_df$lga_name))
```

```{r}
facility_df <- facility_df %>%
  mutate(
    lga_name_raw = lga_name,
    lga_name_source  = str_squish(lga_name),
    lga_name_source  = case_when(
      str_to_lower(lga_name) == "latrobe city council" ~ "Latrobe City Council",
      str_to_lower(lga_name) == "monash city council" ~ "Monash City Council",
      str_to_lower(lga_name) == "mornington peninsula shire council" ~ 
        "Mornington Peninsula Shire Council",
      str_to_lower(lga_name) == "towong shire council" ~ 
        "Towong Shire Council",

      # 2026 LGA
      str_to_lower(lga_name) == "moreland city council" ~ "Merri-bek City Council",
      str_to_lower(lga_name) == "geelong city council" ~ "Greater Geelong City Council",
      str_to_lower(lga_name) == "dandenong city council" ~ "Greater Dandenong City Council",
      str_to_lower(lga_name) == "shepparton city council" ~ "Greater Shepparton City Council",

      TRUE ~ lga_name
    )
  )
```

```{r}
facility_df %>%
  summarise(
    raw_lga = n_distinct(lga_name_raw, na.rm = TRUE),
    n_unique_lga = n_distinct(lga_name_source, na.rm = TRUE),
    missing_lga_records  = sum(is.na(lga_name_source)),
    missing_lga_facilities = n_distinct(
      facility_id_1[is.na(lga_name)]
    )
  )
```

```{r}
facility_df %>%
  filter(is.na(lga_name_source)) %>%
  select(
    facility_id_1,
    facility_name,
    suburb_town,
    pcode,
    latitude,
    longitude
  )
```

Five records had missing LGA values, corresponding to a single facility. Because the facility's address-related fields were inconsistent with its coordinates, the LGA was not manually imputed and will be reassigned using spatial intersection with the 2026 Victorian LGA boundaries.

### 7. Create a Clean Sport-level dataset

There are 9,590 records in the original dataset, corresponding to 5,000 unique facilities and multiple facility-sport/component records. The following code creates a clean sport-level dataset with selected columns.

```{r}
sport_level <- facility_df %>%
  select(
    facility_id = facility_id_1,
    facilty_sport_played_id,
    facility_name,
    lga_name_source,
    latitude,
    longitude,
    facility_ownership,
    facility_purpose,
    facility_category,
    sports_played,
    number_of_field_courts,

    suburb_town,
    field_surface_type
  )

skim(sport_level)
```

### 8. Create a clean facility-level dataset

The following code creates a clean facility-level dataset with one record per facility. The number of distinct sports and total number of fields/courts are calculated for each facility.

```{r}
collapse_unique <- function(x) {
  x <- sort(unique(na.omit(x)))
  
  if (length(x) == 0) {
    NA_character_
  } else {
    paste(x, collapse = "; ")
  }
}

facility_level <- facility_df %>%
    rename(
      facility_id = facility_id_1) %>%
  group_by(facility_id) %>%
  summarise(
    facility_name = first(facility_name),
    lga_name_source = first(lga_name_source),
    latitude = first(latitude),
    longitude = first(longitude),
    facility_ownership = first(na.omit(facility_ownership)),
    facility_purpose = first(na.omit(facility_purpose)),
    facility_category = first(na.omit(facility_category)),
    
    # number of distinct sports recorded at a facility
    n_sports = n_distinct(sports_played, na.rm = TRUE),
    
    sports_played = collapse_unique(sports_played),

    suburb_town = collapse_unique(suburb_town),

    field_surface_type = collapse_unique(field_surface_type),
    
    # total recorded fields/courts
    n_fields_courts = sum(
      number_of_field_courts,
      na.rm = TRUE
    ),
    
    .groups = "drop"
  )

glimpse(facility_level)
skim(facility_level)
dim(facility_level)
n_distinct(facility_level$facility_id)
```

### 9. Spatial Join

Use the 2026 LGA boundaries to assign LGA values to facilities.

```{r}
lga_2026 <- st_read(
  "data/LGA/LGA_POLYGON.shp",
  quiet = TRUE
)

lga_2026
st_crs(lga_2026)
```

```{r}
lga_2026_join <- lga_2026 %>%
  select(
    lga_code = LGA_CODE,
    lga_name_spatial = LGA_NAME,
    geometry
  )
```

#### facility_level 

Spatial join was used as the primary method for assigning LGA based on facility coordinates. For records that could not be assigned through the spatial join, the standardised source LGA was matched to the official 2026 LGA lookup table as a fallback. This fallback was applied both to facilities without valid coordinates and to a small number of coordinate records that did not intersect an LGA polygon.

Of the 5,000 facility records, 4,680 had valid coordinates and were processed using spatial joining. The remaining 320 records without valid coordinates were assigned using the standardised source LGA. Among the 4,680 spatial records, 9 could not be matched to an LGA polygon and were subsequently assigned using the standardised source LGA as a fallback.

After the first join, there are still 2 records with missing LGA codes. 
The source dataset uses the old/official council name, whereas the 2026 spatial LGA dataset uses the standard LGA name.
Bendigo City Council need to be converted to Greater Bendigo City Council, and Borough of Queenscliffe need to be converted to Queenscliffe Council. The following code manually corrects these two records.

```{r}
#  Standardized source LGA
facility_level <- facility_level %>%
  mutate(
    lga_name_standard = lga_name_source %>%
      str_remove(" Rural City Council$") %>%
      str_remove(" Shire Council$") %>%
      str_remove(" City Council$") %>%
      str_remove(" Borough Council$") %>%
      str_to_upper() %>%
      str_trim()
  )
```

```{r}
facility_level <- facility_level %>%
  mutate(
    lga_name_standard = case_when(
      lga_name_standard == "BENDIGO" ~ "GREATER BENDIGO",
      lga_name_standard == "BOROUGH OF QUEENSCLIFFE" ~ "QUEENSCLIFFE",
      TRUE ~ lga_name_standard
    )
  )
```

```{r}
# Set up LGA lookup
lga_lookup <- lga_2026 %>%
  st_drop_geometry() %>%
  transmute(
    lga_code = as.character(LGA_CODE),
    lga_name_spatial = LGA_NAME
  )
```

**4680 spatial join: **

There are 4,680 records with valid coordinates that can be spatially joined to the LGA boundaries.

```{r}
facility_coord_sf <- facility_level %>%
  filter(
    !is.na(longitude),
    !is.na(latitude)
  ) %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )
```

```{r}
lga_2026_join <- lga_2026_join %>%
  st_transform(st_crs(facility_coord_sf))

st_crs(facility_coord_sf)
st_crs(lga_2026_join)
```

```{r}
facility_coord_sf <- facility_coord_sf %>%
  st_join(
    lga_2026_join,
    join = st_intersects,
    left = TRUE
  )
```

```{r}
# Create spatial LGA assignment
facility_coord_lga <- facility_coord_sf %>%
  st_drop_geometry() %>%
  select(
    facility_id,
    lga_code,
    lga_name_spatial
  )
```

```{r}
# Check for any records that still have missing LGA codes after the spatial join
facility_coord_lga %>%
  filter(is.na(lga_code)) %>%
  print(n = Inf)
```

After the spatial join, 4,671 records were successfully assigned to an LGA. Nine records could not be assigned to an LGA through the spatial join.
There are 9 records with valid coordinates that could not be assigned to an LGA through the spatial join. For these nine records, the standardised source LGA was used as a fallback and matched to the official 2026 LGA lookup table.

```{r}
# Identify the 9 records that failed the spatial join
facility_coord_lga_missing <- facility_coord_lga %>%
  filter(is.na(lga_code)) %>%
  select(facility_id)
```

```{r}
facility_coord_lga_fallback <- facility_level %>%
  filter(facility_id %in% facility_coord_lga_missing$facility_id) %>%
  select(
    facility_id,
    lga_name_standard
  ) %>%
  left_join(
    lga_lookup %>%
      rename(
        lga_code_lookup = lga_code,
        lga_name_lookup = lga_name_spatial
      ),
    by = c(
      "lga_name_standard" = "lga_name_lookup"
    )
  ) 
```

```{r}
facility_coord_lga <- facility_coord_lga %>%
  left_join(
    facility_coord_lga_fallback,
    by = "facility_id"
  ) %>%
  mutate(
    lga_code = coalesce(lga_code, lga_code_lookup),
    lga_name_spatial = coalesce(
      lga_name_spatial,
      lga_name_standard
    )
  ) %>%
  select(
    facility_id,
    lga_code,
    lga_name_spatial
  )
```

```{r}
# Check for any records that still have missing LGA codes after the spatial join
facility_coord_lga %>%
  filter(is.na(lga_code)) %>%
  print(n = Inf)
```

**320 Non-spatial fallback: **

There are 320 records without coordinates, which cannot be spatially joined. 
However, we can use the original LGA values to assign LGA codes and names using a lookup table.

```{r}
facility_no_coord_lga <- facility_level %>%
  filter(
    is.na(latitude),
    is.na(longitude)
  ) %>%
  select(
    facility_id,
    lga_name_standard
  ) %>%
  left_join(
    lga_lookup,
    by = c(
      "lga_name_standard" = "lga_name_spatial"
    )
  ) %>%
  select(
    facility_id,
    lga_code,
    lga_name_standard
  ) %>%
  rename(
    lga_name_spatial = lga_name_standard
  )
```


```{r}
# Check for any records that still have missing LGA codes after the non-spatial fallback
facility_no_coord_lga %>%
  filter(is.na(lga_code)) %>%
  print(n = Inf)
```


```{r}
# Merging two LGA assignments
facility_lga <- bind_rows(
  facility_coord_lga,
  facility_no_coord_lga
)
```

```{r}
facility_lga %>%
  summarise(
    rows = n(),
    unique_facilities = n_distinct(facility_id),
    missing_lga_code = sum(is.na(lga_code)),
    missing_lga_name = sum(is.na(lga_name_spatial))
  )
```

```{r}
# Final merge of LGA values into facility_level dataset
facility_level <- facility_level %>%
  select(
    -any_of(c("lga_code", "lga_name_spatial"))
  ) %>%
  left_join(
    facility_lga,
    by = "facility_id"
  )
```

#### sport_level

LGA assignment was first established at the facility level using spatial joins for facilities with valid coordinates and a source-LGA fallback for facilities without coordinates. The resulting LGA assignments were then propagated to the sport-level records using the unique facility ID.

```{r}
sport_level <- sport_level %>%
  left_join(
    facility_level %>%
      select(
        facility_id,
        lga_code,
        lga_name_spatial
      ),
    by = "facility_id"
  )
```

### 9. Spatial Join QA

#### facility_level

```{r}
nrow(facility_level)
n_distinct(facility_level$facility_id)
```

#### sport_level
```{r}
nrow(sport_level)
sport_level %>%
  summarise(
    missing_lga_code = sum(is.na(lga_code)),
    missing_lga_name = sum(is.na(lga_name_spatial))
  )

sport_level %>%
  distinct(facility_id, lga_code, lga_name_spatial) %>%
  count(facility_id) %>%
  filter(n > 1)
```

### 10. Change column names for consistency
```{r}
facility_level <- facility_level %>%
  rename(
    LGA_CODE = lga_code,
    LGA_NAME = lga_name_spatial
  ) %>%
  select(-lga_name_standard)

sport_level <- sport_level %>%
  rename(
    LGA_CODE = lga_code,
    LGA_NAME = lga_name_spatial
  )
```


## Physical activity


### 1.Explore the physical activity data

#### Overview of the physical activity data

```{r}
dim(activity)
head(activity)
names(activity)
glimpse(activity)
```

#### Check available physical activity indicators

```{r}
# Display the available physical activity indicators
activity |>
  distinct(
    Indicator,
    `Indicator category`
  ) |>
  arrange(
    Indicator,
    `Indicator category`
  )

```

```{r}
# Check the available estimate types and gender categories
activity |>
  count(
    `Estimate type`,
    Gender,
    sort = TRUE
  )
```
### 2. Select the physical activity outcome

#### Extract LGA-level observations


```{r}
# Retain observations reported for LGAs within each
# Local Public Health Unit
activity_lga_all <- activity |>
  filter(
    str_starts(`Stratified by`, "LGAs of ")
  )
nrow(activity_lga_all)
```

#### Select the age-adjusted physical activity indicator

```{r}
# Select the age-adjusted percentage of people who complete
# at least 150 minutes of moderate-to-vigorous
# physical activity per week
activity_lga_selected <- activity_lga_all |>
  filter(
    Indicator == "1. Moderate to vigourous physical activity",
    `Indicator category` == "≥150 minutes/week",
    `Estimate type` == "Age-adjusted",
    Gender == "People"
  )
```

#### Remove Victoria-wide summary records

```{r}
# Remove the repeated Victoria-wide estimate
# Each PHU section contains the same Victoria summary row
activity_lga_selected <- activity_lga_selected |>
  filter(
    `Sub-population` != "Victoria"
  )
```

