---
title: "Recreational Spaces and Health Outcomes Across Victoria"
format: html
editor: visual
---

```{r}
library(readxl)
library(sf)
library(dplyr)
library(ggplot2)
library(stringr)
```

## Load data

```{r}
# Load the Parks and Conservation Reserves spatial dataset
parks <- st_read("data/parks/PARKRES.shp")

# Load the sport and recreational facilities dataset
facilities <- read_excel("data/facilities.xlsx")

# Load LGA-level physical activity and health data
# n_max is used to exclude non-data rows at the bottom of the worksheet
activity <- read_excel(
  "data/Victorian Population Health Survey 2023.xlsx",
  n_max = 8910
)

# Load Victorian Local Government Area boundaries
lga <- st_read("data/LGA/LGA_POLYGON.shp")
```

## parks

### Overview of parks across Victorian LGAs

The following map shows all parks and conservation reserves in the original dataset. It provides an initial overview of their spatial distribution across Victorian LGAs before recreational park types are selected.

```{r, fig.width=8, fig.height=7}
# Map all parks and conservation reserves together with LGA boundaries
ggplot() +
  geom_sf(
    data = parks,
    fill = "#63A65F",
    colour = NA,
    alpha = 0.75
  ) +
  geom_sf(
    data = lga,
    fill = NA,
    colour = "grey30",
    linewidth = 0.25
  ) +
  labs(
    title = str_wrap(
      "Distribution of All Parks and Conservation Reserves Across Victorian LGAs",
      width = 55
    ),
    subtitle = "LGA boundaries are shown in grey",
    caption = paste(
      "Source: Victorian Parks and Conservation",
      "Reserves and LGA boundary datasets"
    )
  ) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold",
      lineheight = 1.1
    ),
    plot.subtitle = element_text(
      size = 11,
      colour = "grey40"
    ),
    plot.caption = element_text(
      size = 9,
      colour = "grey50"
    ),
    plot.margin = margin(15, 20, 15, 20)
  )
```


### Inspect park data structure

```{r,echo=FALSE}
# Examine the variables and data types in the park dataset
glimpse(parks)
```


### Check key variables

```{r}
# Check missing values in key park variables
parks |>
  st_drop_geometry() |>
  summarise(
    missing_id = sum(is.na(PRIMS_ID)),
    missing_name = sum(is.na(NAME)),
    missing_area_type = sum(is.na(AREA_TYPE))
  )
```

### Check park geometries

```{r}
# Check for empty and invalid park geometries
parks |>
  summarise(
    empty_geometries = sum(st_is_empty(geometry)),
    invalid_geometries = sum(!st_is_valid(geometry))
  ) |>
  st_drop_geometry()
```

### Check park identifiers

```{r}
# Compare the number of polygon records with unique park identifiers
parks |>
  st_drop_geometry() |>
  summarise(
    polygon_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID),
    additional_polygon_records = n() - n_distinct(PRIMS_ID)
  )
```

### Check park attribute consistency

```{r}
# Check whether each park ID has consistent names and classifications
parks |>
  st_drop_geometry() |>
  group_by(PRIMS_ID) |>
  summarise(
    n_names = n_distinct(NAME),
    n_area_types = n_distinct(AREA_TYPE),
    .groups = "drop"
  ) |>
  filter(n_names > 1 | n_area_types > 1)
```


#### Inspect inconsistent park attributes

```{r}
# Display the names and classifications of inconsistent park IDs
parks |>
  st_drop_geometry() |>
  filter(PRIMS_ID %in% c(18, 694, 4654, 5077, 5089)) |>
  distinct(PRIMS_ID, NAME, AREA_TYPE) |>
  arrange(PRIMS_ID, NAME) |>
  print(n = Inf)
```

#### Examine inconsistent park records

```{r}
# Examine additional attributes of inconsistent park records
parks |>
  st_drop_geometry() |>
  filter(PRIMS_ID %in% c(18, 694, 4654, 5077, 5089)) |>
  select(
    PRIMS_ID,
    NAME,
    AREA_TYPE,
    AREATYPEID,
    ESTAB_DATE,
    LAST_MOD,
    HECTARES
  ) |>
  arrange(PRIMS_ID, NAME, LAST_MOD) |>
  print(n = Inf)
```


The checks identified no missing values or invalid geometries. Multiple records
with the same `PRIMS_ID` mainly represent separate components or subdivisions
of the same park, so these records are retained.

### Inspect park classifications

```{r}
# Count unique park IDs in each park classification
parks |>
  st_drop_geometry() |>
  distinct(PRIMS_ID, AREA_TYPE) |>
  count(AREA_TYPE, sort = TRUE) |>
  print(n = Inf)
```
### Select recreational park types

```{r}
# Retain classifications that are clearly recreational parks
recreational_types <- c(
  "COASTAL RESERVE",
  "FOREST PARK",
  "METROPOLITAN PARK",
  "NATIONAL PARK - SCHEDULE 2, NATIONAL PARKS ACT",
  "NATIONAL PARKS ACT SCHEDULE 4 PARK OR RESERVE",
  "OTHER PARK - SCHEDULE 3, NATIONAL PARKS ACT",
  "OTHER PARK - SCHEDULE 9, NATIONAL PARKS ACT",
  "PHILLIP ISLAND NATURE PARK",
  "REGIONAL PARK - NOT SCHEDULED UNDER NATIONAL PARKS ACT",
  "RESERVOIR PARK",
  "STATE PARK - SCHEDULE 2B, NATIONAL PARKS ACT",
  "WILDERNESS PARK - SCHEDULE 2A, NATIONAL PARKS ACT"
)

parks_selected <- parks |>
  filter(AREA_TYPE %in% recreational_types)
```

### Check selected recreational parks

```{r}
# Compare the original and selected park datasets
bind_rows(
  parks |>
    st_drop_geometry() |>
    summarise(
      dataset = "All park records",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    ),
  parks_selected |>
    st_drop_geometry() |>
    summarise(
      dataset = "Selected recreational parks",
      polygon_records = n(),
      unique_park_ids = n_distinct(PRIMS_ID)
    )
)
```


### Summarise selected park types

```{r}
# Count selected park IDs by classification
parks_selected |>
  st_drop_geometry() |>
  distinct(PRIMS_ID, AREA_TYPE) |>
  count(AREA_TYPE, sort = TRUE) |>
  print(n = Inf)
```

### Combine park polygons

```{r}
# Combine multiple polygon components belonging to the same park
parks_dissolved <- parks_selected |>
  group_by(PRIMS_ID) |>
  summarise(
    NAME = first(NAME),
    AREA_TYPE = first(AREA_TYPE),
    .groups = "drop"
  )
```

```{r}
# Confirm that each park ID now appears once
parks_dissolved |>
  st_drop_geometry() |>
  summarise(
    park_records = n(),
    unique_park_ids = n_distinct(PRIMS_ID)
  )
```

### Check coordinate reference systems

```{r}
# Check whether parks and LGA boundaries use the same CRS
st_crs(parks_dissolved) == st_crs(lga)
```


### Assign parks to LGAs for park counts

For park-count calculations, each park is assigned to the LGA containing its
largest overlapping area. This ensures that parks crossing multiple LGAs are
counted only once.

```{r}
# Assign each park to the LGA with the largest overlapping area
parks_lga_count <- st_join(
  parks_dissolved,
  lga |>
    select(LGA_CODE, LGA_NAME),
  left = TRUE,
  largest = TRUE
)
```

```{r}
sum(is.na(parks_lga_count$LGA_NAME))
```

### Inspect unmatched parks

```{r}
# Display parks that were not assigned to an LGA
parks_lga_count |>
  filter(is.na(LGA_NAME)) |>
  st_drop_geometry() |>
  select(PRIMS_ID, NAME, AREA_TYPE)
```


### Check nearest LGAs for unmatched parks

```{r}
# Extract parks that do not intersect an LGA
unmatched_parks <- parks_lga_count |>
  filter(is.na(LGA_NAME))

# Identify the nearest LGA to each unmatched park
nearest_lga_index <- st_nearest_feature(unmatched_parks, lga)

tibble(
  PRIMS_ID = unmatched_parks$PRIMS_ID,
  NAME = unmatched_parks$NAME,
  nearest_LGA = lga$LGA_NAME[nearest_lga_index],
  distance_metres = as.numeric(
    st_distance(
      unmatched_parks,
      lga[nearest_lga_index, ],
      by_element = TRUE
    )
  )
)
```












