---
title: "EDA"
format: html
---

1. Load packages and data

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(GGally)
library(scales)
```

```{r}
accessibility_raw_lga <- read_csv(
  "clean_data/accessibility_raw_lga.csv"
)

health_lga <- read_csv(
  "clean_data/health_lga.csv"
)

lga <- st_read(
  "data/lga/LGA_POLYGON.shp",
  quiet = TRUE
)
```

2. Prepare analysis dataset
```{r}
# Check LGA names
accessibility_raw_lga |>
  summarise(
    n_lgas = n(),
    n_unique_lgas = n_distinct(LGA_CODE)
  )

health_lga |>
  summarise(
    n_lgas = n(),
    n_unique_lgas = n_distinct(LGA_NAME)
  )

analysis_lga <- accessibility_raw_lga |>
  left_join(
    health_lga,
    by = "LGA_NAME"
  )
```

```{r}
analysis_lga |>
  summarise(
    n_lgas = n(),
    missing_health = sum(is.na(poor_health_percent)),
    missing_obesity = sum(is.na(obesity_percent)),
    missing_distress = sum(is.na(distress_percent))
  )
```

3. Basic descriptive statistics
```{r}
accessibility_vars <- c(
  "park_count",
  "park_area_hectares",
  "facility_count",
  "total_fields_courts",
  "mean_sports_per_facility",
  "sport_opportunity_count",
  "sport_type_count",
  "facilities_with_sport_data"
)

accessibility_summary <- analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  group_by(variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

accessibility_summary
```



4. Distribution of accessibility indicators

```{r}
analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(
    bins = 20
  ) +
  facet_wrap(
    ~ variable,
    scales = "free",
    ncol = 2
  ) +
  labs(
    title = "Distribution of Recreational Accessibility Indicators",
    x = NULL,
    y = "Number of LGAs"
  ) +
  theme_minimal()
```

5. Boxplots and outliers
```{r}
analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Variation in Recreational Accessibility Indicators",
    x = NULL,
    y = "Value"
  ) +
  theme_minimal()
```

6. Identify extreme LGAs
```{r}
analysis_lga |>
  arrange(desc(park_area_hectares)) |>
  select(
    LGA_NAME,
    park_area_hectares
  ) |>
  slice_head(n = 10)
```

```{r}
analysis_lga |>
  arrange(desc(facility_count)) |>
  select(
    LGA_NAME,
    facility_count
  ) |>
  slice_head(n = 10)
```

```{r}
analysis_lga |>
  arrange(desc(sport_opportunity_count)) |>
  select(
    LGA_NAME,
    sport_opportunity_count
  ) |>
  slice_head(n = 10)
```

7. Correlation between accessibility indicators
```{r}
accessibility_cor <- analysis_lga |>
  select(all_of(accessibility_vars)) |>
  cor(
    use = "pairwise.complete.obs",
    method = "pearson"
  )

round(accessibility_cor, 2)
```

correlation matrix：
```{r}
GGally::ggcorr(
  analysis_lga |>
    select(all_of(accessibility_vars)),
  label = TRUE,
  label_round = 2,
  hjust = 0.8,
  layout.exp = 2
)
```

8. Spearman correlation
Many accessibility variables are likely skewed and there are only 79 LGAs, Spearman's correlation is also checked.

```{r}
accessibility_spearman <- analysis_lga |>
  select(all_of(accessibility_vars)) |>
  cor(
    use = "pairwise.complete.obs",
    method = "spearman"
  )

round(accessibility_spearman, 2)
```

9. Scatter plots of key accessibility indicators
```{r}
ggplot(
  analysis_lga,
  aes(
    x = facility_count,
    y = sport_opportunity_count
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Facility Count and Sport Opportunities",
    x = "Number of Facilities",
    y = "Number of Sport Opportunities"
  ) +
  theme_minimal()
```


```{r}
ggplot(
  analysis_lga,
  aes(
    x = park_count,
    y = park_area_hectares
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Park Count and Park Area",
    x = "Number of Parks",
    y = "Park Area (hectares)"
  ) +
  theme_minimal()
```

10. Spatial EDA

```{r}
lga_map <- lga |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  mutate(LGA_CODE = as.numeric(LGA_CODE)) |>   # 加这一行
  left_join(
    analysis_lga,
    by = c("LGA_CODE", "LGA_NAME")
  )
```

10.1 Park count map

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = park_count),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Park count"
  ) +
  labs(
    title = "Number of Parks across Victorian LGAs"
  ) +
  theme_minimal()

```

10.2 Facility count map

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = facility_count),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Facilities"
  ) +
  labs(
    title = "Number of Recreational Facilities across Victorian LGAs"
  ) +
  theme_minimal()
```

10.3 Sport opportunity map

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = sport_opportunity_count),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Sport opportunities"
  ) +
  labs(
    title = "Sport Opportunities across Victorian LGAs"
  ) +
  theme_minimal()
```