---
title: "EDA"
format: html
---

1. Load packages and data

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(GGally)
library(scales)
library(corrplot)
```

```{r}
accessibility_lga <- read_csv(
  "clean_data/accessibility_lga.csv"
)

health_lga <- read_csv(
  "clean_data/health_lga.csv"
)

physical_activity_lga <- read_csv(
  "clean_data/physical_activity_lga.csv"
)

lga <- st_read(
  "data/lga/LGA_POLYGON.shp",
  quiet = TRUE
)

seifa_lga <- read_csv("clean_data/seifa_lga.csv")
```

2. Prepare analysis dataset
```{r}
# Check LGA names
accessibility_lga |>
  summarise(
    n_lgas = n(),
    n_unique_lgas = n_distinct(LGA_CODE)
  )

health_lga |>
  summarise(
    n_lgas = n(),
    n_unique_lgas = n_distinct(LGA_NAME)
  )

physical_activity_lga |>
  summarise(
    n_lgas = n(),
    n_unique_lgas = n_distinct(LGA_NAME)
  )

analysis_lga <- accessibility_lga |>
  left_join(
    health_lga,
    by = "LGA_NAME"
  )

analysis_lga <- analysis_lga |>
  left_join(
    physical_activity_lga |>
      select(
        -LGA_NAME_ORIGINAL
      ),
    by = "LGA_NAME"
  )
```

```{r}
analysis_lga |>
  summarise(
    n_lgas = n(),
    missing_health = sum(is.na(poor_health_percent)),
    missing_obesity = sum(is.na(obesity_percent)),
    missing_distress = sum(is.na(distress_percent)),
    missing_activity = sum(is.na(activity_percent)),
    missing_lower_ci = sum(is.na(lower_ci)),
    missing_upper_ci = sum(is.na(upper_ci))
  )
```

3. Basic descriptive statistics
```{r}
accessibility_vars <- c(
  "parks_per_10000",
  "park_area_sqm_per_capita",
  "facilities_per_10000",
  "fields_courts_per_10000",
  "sport_opportunities_per_10000",
  "sport_type_count",
  "mean_sports_per_facility"
)

accessibility_summary <- analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  group_by(variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

accessibility_summary
```

```{r}
activity_summary <- analysis_lga |>
  summarise(
    n = sum(!is.na(activity_percent)),
    mean = mean(activity_percent, na.rm = TRUE),
    median = median(activity_percent, na.rm = TRUE),
    sd = sd(activity_percent, na.rm = TRUE),
    min = min(activity_percent, na.rm = TRUE),
    q1 = quantile(activity_percent, 0.25, na.rm = TRUE),
    q3 = quantile(activity_percent, 0.75, na.rm = TRUE),
    max = max(activity_percent, na.rm = TRUE)
  )

activity_summary
```

4. Distribution of accessibility indicators

```{r}
analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(
    bins = 20
  ) +
  facet_wrap(
    ~ variable,
    scales = "free",
    ncol = 2
  ) +
  labs(
    title = "Distribution of Recreational Accessibility Indicators",
    x = NULL,
    y = "Number of LGAs"
  ) +
  theme_minimal()
```

Several accessibility indicators were strongly right-skewed. In particular, parks per 10,000 residents and park area per capita had large positive skewness, while facility-related rate measures were also moderately to strongly right-skewed. In contrast, sport type count and mean sports per facility were approximately symmetric.

5. Distribution of physical activity
```{r}
ggplot(
  analysis_lga,
  aes(x = activity_percent)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Physical Activity across Victorian LGAs",
    x = "Physical activity (%)",
    y = "Number of LGAs"
  ) +
  theme_minimal()
```

6. Boxplots and outliers
```{r}
analysis_lga |>
  select(all_of(accessibility_vars)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Variation in Recreational Accessibility Indicators",
    x = NULL,
    y = "Value"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  analysis_lga,
  aes(y = activity_percent)
) +
  geom_boxplot() +
  labs(
    title = "Variation in Physical Activity across Victorian LGAs",
    y = "Physical activity (%)"
  ) +
  theme_minimal()
```

7. Identify extreme LGAs
```{r}
analysis_lga |>
  select(
    LGA_NAME,
    all_of(accessibility_vars)
  ) |>
  arrange(desc(parks_per_10000)) |>
  slice_head(n = 10)
```

```{r}
analysis_lga |>
  select(
    LGA_NAME,
    all_of(accessibility_vars)
  ) |>
  arrange(desc(facilities_per_10000)) |>
  slice_head(n = 10)
```

8. Correlation between accessibility indicators
```{r}
accessibility_cor <- accessibility_lga |>
  select(
    all_of(accessibility_vars)
  ) |>
  cor(
    use = "pairwise.complete.obs",
    method = "spearman"
  ) |>
  round(2)

# correlation matrix
accessibility_cor

# scatterplot matrix
GGally::ggpairs(
  accessibility_lga |>
    select(
      all_of(accessibility_vars)
    )
) +
  theme(
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 6)
  )

# correlation heatmap
corrplot(
  accessibility_cor,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.7,
  number.cex = 0.7
)
```

Strong correlations were observed among several accessibility measures. The three facility-related measures were especially highly correlated, with correlations ranging from 0.94 to 0.99. The two park measures were also strongly correlated. These highly correlated variables should not all be included in the same regression model.


9. Scatter plots of key accessibility indicators
```{r}
ggplot(
  analysis_lga,
  aes(
    x = facilities_per_10000,
    y = sport_opportunities_per_10000
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Facility Count and Sport Opportunities",
    x = "Facilities per 10,000 residents",
    y = "Sport opportunities per 10,000 residents"
  ) +
  theme_minimal()
```


```{r}
ggplot(
  analysis_lga,
  aes(
    x = parks_per_10000,
    y = park_area_sqm_per_capita
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Park Count and Park Area",
    x = "Parks per 10,000 residents",
    y = "Park area (sqm per capita)"
  ) +
  theme_minimal()
```

10. Spatial EDA

```{r}
lga_map <- lga |>
  filter(!str_detect(LGA_NAME, "UNINC")) |>
  mutate(LGA_CODE = as.numeric(LGA_CODE)) |>   
  left_join(
    analysis_lga,
    by = c("LGA_CODE", "LGA_NAME")
  )
```

10.1 Park provision

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = parks_per_10000),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Parks per 10,000"
  ) +
  labs(
    title = "Park Provision across Victorian LGAs"
  ) +
  theme_void()

```

10.2 Park area
```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = park_area_sqm_per_capita),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Park area per capita"
  ) +
  labs(
    title = "Park Area per Capita across Victorian LGAs"
  ) +
  theme_void()
```


10.3 Facility provision

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = facilities_per_10000),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Facilities per 10,000"
  ) +
  labs(
    title = "Facility Provision across Victorian LGAs"
  ) +
  theme_void()
```

10.4 Sport opportunity map

```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = sport_opportunities_per_10000),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Sport opportunities per 10,000"
  ) +
  labs(
    title = "Sport Opportunity Provision across Victorian LGAs"
  ) +
  theme_void()
```

10.5 Physical activity map
```{r}
ggplot(lga_map) +
  geom_sf(
    aes(fill = activity_percent),
    colour = NA
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Physical activity (%)"
  ) +
  labs(
    title = "Physical Activity across Victorian LGAs"
  ) +
  theme_void()
```

11. Health outcome EDA
```{r}
health_vars <- c(
  "obesity_percent",
  "distress_percent",
  "poor_health_percent"
)
```

```{r}
analysis_lga |>
  select(all_of(health_vars)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_wrap(
    ~ variable,
    scales = "free",
    ncol = 1
  ) +
  labs(
    title = "Distribution of Health Outcomes",
    x = "Percentage",
    y = "Number of LGAs"
  ) +
  theme_minimal()
```


12. Accessibility vs health outcomes
```{r}
outcome_vars <- c(
  "poor_health_percent",
  "obesity_percent",
  "distress_percent"
)

accessibility_outcome_cor <- analysis_lga |>
  select(
    all_of(accessibility_vars),
    all_of(outcome_vars)
  ) |>
  cor(
    use = "pairwise.complete.obs",
    method = "spearman"
  ) |>
  round(2)

accessibility_outcome_cor
```


```{r}
ggplot(
  analysis_lga,
  aes(
    x = parks_per_10000,
    y = obesity_percent
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "#B2182B") +
  labs(
    title = "Parks per 10,000 residents and Obesity",
    x = "Parks per 10,000 residents",
    y = "Obesity (%)"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  analysis_lga,
  aes(
    x = park_area_sqm_per_capita,
    y = obesity_percent
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "#B2182B") +
  labs(
    title = "Park Area per Capita and Obesity",
    x = "Park area (sqm per capita)",
    y = "Obesity (%)"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  analysis_lga,
  aes(
    x = parks_per_10000,
    y = poor_health_percent
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "#B2182B") +
  labs(
    title = "Parks per 10,000 residents and Poor Health",
    subtitle = "Spearman correlation = 0.38",
    x = "Parks per 10,000 residents",
    y = "Poor Health (%)"
  ) +
  theme_minimal()
```

Several accessibility indicators were positively correlated with adverse health outcomes. In particular, parks per 10,000 residents showed positive correlations with obesity (0.60), followed by poor health (0.38) and psychological distress (0.25). However, sport type count and mean sports per facility showed much weaker relationships with health outcomes. These findings suggest that the relationships vary across different accessibility measures, and greater recreational provision per capita is not necessarily associated with better health outcomes at the LGA level. The scatterplots also show considerable variation around the fitted regression lines, suggesting that accessibility alone does not adequately explain differences in health outcomes across LGAs.

13. Accessibility vs Physical Activity
```{r}
activity_cor <- analysis_lga |>
  select(
    all_of(accessibility_vars),
    activity_percent
  ) |>
  cor(
    use = "pairwise.complete.obs",
    method = "spearman"
  ) |>
  round(2)

activity_cor
```

Accessibility indicators showed weak negative correlations with physical activity, ranging from -0.24 to -0.02. Mean sports per facility had the strongest negative correlation (-0.24), while facilities per 10,000 residents showed almost no correlation (-0.02). Overall, these results suggest that recreational accessibility indicators alone have limited associations with physical activity at the LGA level.

```{r}
ggplot(analysis_lga, aes(park_area_sqm_per_capita, activity_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Park Area vs Physical Activity",
       x = "Park area per capita", y = "Activity (%)") +
  theme_minimal()

# parks
ggplot(analysis_lga, aes(parks_per_10000, activity_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Parks vs Physical Activity",
       x = "Parks per 10,000", y = "Activity (%)") +
  theme_minimal()

ggplot(analysis_lga, aes(fields_courts_per_10000, activity_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Fields/Courts vs Physical Activity",
       x = "Fields/courts per 10,000", y = "Activity (%)") +
  theme_minimal()
```

14. Physical activity vs Health outcome
```{r}
activity_health_cor <- analysis_lga |>
  select(
    activity_percent,
    poor_health_percent,
    obesity_percent,
    distress_percent
  ) |>
  cor(
    use = "pairwise.complete.obs",
    method = "spearman"
  ) |>
  round(2)

activity_health_cor
```

```{r}
ggplot(analysis_lga, aes(activity_percent, obesity_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Activity vs Obesity",
       x = "Activity (%)", y = "Obesity (%)") +
  theme_minimal()

ggplot(analysis_lga, aes(activity_percent, poor_health_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Activity vs Poor Health",
       x = "Activity (%)", y = "Poor Health (%)") +
  theme_minimal()

ggplot(analysis_lga, aes(activity_percent, distress_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Activity vs Distress",
       x = "Activity (%)", y = "Distress (%)") +
  theme_minimal()
```

Physical activity is negatively correlated with all three health outcomes: obesity (-0.41), poor health (-0.41), and distress (-0.20). This suggests higher activity is associated with better health, though the scatterplots show a moderate fit rather than a strong linear relationship.


15. Socioeconomic Status (SEIFA)

15.1 SEIFA vs Physical Activity and Health Outcomes

```{r}
# Join SEIFA data
analysis_lga <- analysis_lga |>
  left_join(
    seifa_lga |>
      select(LGA_NAME, SEIFA_IRSAD),
    by = "LGA_NAME"
  )

# Correlations between SEIFA and outcomes
seifa_outcome_cor <- analysis_lga |>
  select(
    SEIFA_IRSAD,
    activity_percent,
    obesity_percent,
    distress_percent,
    poor_health_percent
  ) |>
  cor(
    method = "spearman",
    use = "pairwise.complete.obs"
  ) |>
  round(2)

seifa_outcome_cor

analysis_lga |>
  select(
    SEIFA_IRSAD,
    activity_percent,
    obesity_percent,
    distress_percent,
    poor_health_percent
  ) |>
  pivot_longer(
    cols = -SEIFA_IRSAD,
    names_to = "outcome",
    values_to = "value"
  ) |>
  ggplot(aes(x = SEIFA_IRSAD, y = value)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title = "Socioeconomic Status and Health Outcomes",
    x = "SEIFA IRSAD Score",
    y = "Percentage (%)"
  ) +
  theme_minimal()
```

SEIFA IRSAD was positively correlated with physical activity (0.35) and negatively correlated with obesity (-0.73), poor health (-0.66), and psychological distress (-0.51). This suggests that more socioeconomically advantaged LGAs tend to have higher physical activity levels and better health outcomes. 

15.2 SEIFA vs Accessibility

```{r}
seifa_accessibility_cor <- analysis_lga |>
  select(
    SEIFA_IRSAD,
    all_of(accessibility_vars)
  ) |>
  cor(
    method = "spearman",
    use = "pairwise.complete.obs"
  ) |>
  round(2)

seifa_accessibility_cor
```

SEIFA IRSAD was negatively correlated with all five per-capita recreational provision indicators, with correlations ranging from -0.65 to -0.53. This indicates that more socioeconomically advantaged LGAs tend to have fewer recreational resources per resident. In contrast, sport type count and mean sports per facility showed little association with SEIFA. These findings suggest that socioeconomic differences may contribute to the observed relationships between recreational provision and health outcomes.



## Create master dataset

```{r}
master_lga <- accessibility_lga |>
  left_join(
    physical_activity_lga |>
      select(-LGA_NAME_ORIGINAL),
    by = "LGA_NAME"
  ) |>
  left_join(
    health_lga,
    by = "LGA_NAME"
  ) |>
  left_join(
    seifa_lga |>
      select(LGA_NAME, SEIFA_IRSAD),
    by = "LGA_NAME"
  )

# Save master dataset
write_csv(master_lga, "clean_data/master_lga.csv")
```