---
title: "Model regression"
format: html
---

```{r}
library(tidyverse)
library(car)
library(broom)
```

```{r}
master_lga <- read_csv("clean_data/master_lga.csv")
```

## 1. Model specification

Outcomes: physical activity, obesity, poor health, distress

Main X: parks_per_10000, facilities_per_10000, sport_type_count

Control variable: SEIFA, population density


## 2. Regression analysis

### 2.1 Accessibility and Physical Activity

#### Unadjusted models

```{r}
# Park
m_park_unadj <- lm(
  activity_percent ~ log1p(parks_per_10000),
  data = master_lga
)

# Facility
m_facility_unadj <- lm(
  activity_percent ~ log1p(facilities_per_10000),
  data = master_lga
)

# Sport diversity
m_sport_unadj <- lm(
  activity_percent ~ sport_type_count,
  data = master_lga
)
```

#### SEIFA-adjusted models

```{r}
# Park
m_park <- update(
  m_park_unadj,
  . ~ . + SEIFA_IRSAD
)

# Facility
m_facility <- update(
  m_facility_unadj,
  . ~ . + SEIFA_IRSAD
)

# Sport diversity
m_sport <- update(
  m_sport_unadj,
  . ~ . + SEIFA_IRSAD
)
```

#### Joint accessibility models

```{r}
# Park + Facility + SEIFA
m_joint <- lm(
  activity_percent ~
    log1p(parks_per_10000) +
    log1p(facilities_per_10000) +
    SEIFA_IRSAD,
  data = master_lga
)

# Park + Facility + Sport diversity + SEIFA
m_full <- update(
  m_joint,
  . ~ . + sport_type_count
)
```

#### Population density 

```{r}
m_park_density <- update(
  m_park,
  . ~ . + log1p(population_density_2023)
)

m_facility_density <- update(
  m_facility,
  . ~ . + log1p(population_density_2023)
)

m_sport_density <- update(
  m_sport,
  . ~ . + log1p(population_density_2023)
)

m_joint_density <- update(
  m_joint,
  . ~ . + log1p(population_density_2023)
)

m_full_density <- update(
  m_full,
  . ~ . + log1p(population_density_2023)
)
```


### 2.2 Model Comparison

#### VIF checks

```{r}
activity_models <- list(
  park_unadjusted = m_park_unadj,
  park = m_park,
  park_density = m_park_density,

  facility_unadjusted = m_facility_unadj,
  facility = m_facility,
  facility_density = m_facility_density,

  sport_unadjusted = m_sport_unadj,
  sport = m_sport,
  sport_density = m_sport_density,

  joint = m_joint,
  joint_density = m_joint_density,

  full = m_full,
  full_density = m_full_density
)

vif_results <- lapply(
  activity_models,
  \(m) {
    if (length(coef(m)) <= 2) {
      return(NA)
    }
    car::vif(m)
  }
)

vif_results
```

Using a VIF threshold of 10, the joint model including park provision, facility provision, sport diversity and SEIFA shows no severe multicollinearity. However, adding population density substantially increases VIF values in models containing park provision, suggesting there are notable overlaps between park provision and population density.

Therefore, the joint model without density is retained as a candidate main specification.

#### Nested model comparison

```{r}
# Adding SEIFA
anova(m_park_unadj, m_park)
anova(m_facility_unadj, m_facility)
anova(m_sport_unadj, m_sport)
```

For all three separate accessibility models, adding SEIFA significantly improves their model fit (p < 0.001), indicating that socioeconomic differences explains some variations in physical activity across LGAs.

```{r}
# Adding accessibility dimensions
anova(m_park, m_joint)       # Add facility
anova(m_facility, m_joint)   # Add park
anova(m_joint, m_full)       # Add sport diversity
```

Adding facility provision to a model containing park provision and SEIFA significantly improves model fit (F = 4.87, p = 0.030). In contrast, adding park provision to the facility model does not significantly improve fit (F = 1.59, p = 0.211). Adding sport diversity to the joint park–facility model provides little additional explanatory information (F = 0.14, p = 0.705).


```{r}
# Adding population density
anova(m_park, m_park_density)
anova(m_facility, m_facility_density)
anova(m_sport, m_sport_density)
anova(m_joint, m_joint_density)
anova(m_full, m_full_density)
```

### 2.3 Regression Summary

```{r}
activity_results <- purrr::imap_dfr(
  activity_models,
  \(m, name) {
    broom::tidy(m, conf.int = TRUE) |>
      mutate(model = name, .before = 1)
  }
)

activity_results |>
  filter(term != "(Intercept)") |>
  select(model, term, estimate, conf.low, conf.high, p.value)
```

### 2.4 Diagnotics